The error in Bayesian phylogenetic reconstruction when speciation co-occurs

### 2019-08-22 razzo-methods.tex
Do not use '\input' in file that is already added by an '\input' as it does not scale (that is, this TeX cannot be used as part of a thesis)
Improve wording, added to describe the nested sampling

In the MBD model, parameters $\lambda$ and $\mu$ correspond, respectively, to the common per-species speciation and extinction rates already present in the standard BD model. 
Additionally, MBD relies on two additional parameters, $\nu$ and $q$. 
The first, $\nu$, is the rate at which an environmental change is triggered.
When such event is triggered, each species present in the phylogeny at that moment has a probability $q$ to speciate at that time.
This kind of speciation is of a different nature respect to the one triggered by $\lambda$. In fact, whereas parameter $\lambda$ can be seen as describing a sympatric process, $\nu$ induces the rise of geographical barriers interrupting the gene flow \giovanni{@richel: please check if biology here is accurate} and leading to an allopatric speciation.
Even though multiple speciations can co-occur, polytomies are not allowed in such process as each species can speciate only once at the time.
A likelihood expression for the process is provided in \cite{mbd}.

@@

We simulate the speciation process in continuous time using the Doob-Gillespie algorithm, using the \verb;mbd_sim; function from the \verb;mbd; package (\citep{mbd}).
We let parameters vary using all possible combinations of values as shown in Table~\ref{tab:simulation_parameters}.
For each parameter setting, we generate $1000$ independent reconstructed trees of the same crown age.
\giovanni{Do we need to add more information on the Doob-Gillespie algorithm or it is overkilling?}

\input{razzo-tables/tab_simparameters.tex}

@@

From each MBD tree, we measure the impact of using the simpler BD tree prior on the inference, using the \verb;pirouette; R package, as described in detail in \citep{pirouette}.

In brief: \verb;pirouette; starts from a 'true' (but unobservable in nature) starting phylogeny, from which a DNA sequence alignment (which is observable in nature) is simulated. 
From each sequence alignment, a Bayesian inference is run, to obtain a posterior distribution of jointly-estimated trees and model parameter estimates.
By comparing the true tree and the posterior trees, an inference error distribution is generated. 
We use the twinning option available in \verb;pirouette; to measure a minimum and full error, in which the minimum error is caused by stochasticity in the full pipeline, where the full error is the added error from using an invalid but standard speciation model \giovanni{I am not sure I get this. What do you mean?}.

In our context, the alignments are 1000 nucleotides in length, with a known root sequence of four 250 mono-nucleotide blocks, following the simplest nucleotide substitution model (JC69) and clock model (strict), with a mutation rate of $\frac{1}{2}\cdot c$,  in which $c$ is the crown age. With this mutation rate, each nucleotide is expected to mutate (both silently and non-silently) 
in half of the histories from the root sequence in the past to the sequence in the tips in the present.

For the Bayesian inference, we use a JC69 site model, a strict
clock model and a BD tree prior. Additionally, we use 
an MRCA prior equal to the crown age with a normal distribution 
of width $\sigma = 0.01$. We pick an MCMC setup of 10M states,
which is sampled each 1k moves.

For the error measurement, we use the nLTT statistic \citet{janzen2015} and
a burn-in fraction of $10\%$.

For the twinning, we let the twin trees follow a BD model.

For the experiments, we use both a hand-picked generative model
and a set of candidate models. We set a BD tree prior, JC69 site
model and strict clock model as the generative model. We used
all other combinations of four tree priors, two clock models and
five speciation models, resulting in a set of 39 candidate models.

>>>>

In the MBD model, parameters $\lambda$ and $\mu$ correspond, respectively, 
to the common per-species speciation and extinction rates 
already present in the standard BD model. 
Additionally, MBD relies on two additional parameters, $\nu$ and $q$. 
The first, $\nu$, is the rate at which an environmental change is triggered.
When such event is triggered, each species present in the phylogeny 
at that moment has a probability $q$ to speciate at that time.
This kind of speciation is of a different nature with respect to 
the one triggered by $\lambda$. 
In fact, whereas parameter $\lambda$ can be seen as 
describing a sympatric process, $\nu$ induces the rise of 
geographical barriers that interrupt the gene flow,
resulting in allopatric speciation. 
Even though multiple speciations can co-occur, 
polytomies are not allowed in such process as each species can speciate only 
once during a co-occuring speciation event.
A likelihood expression for the process is provided in \cite{mbd}.

@@

We simulate the speciation process in continuous time 
using the \verb;mbd_sim; function from the \verb;mbd; R package (\citep{mbd}).
We let parameters vary using all possible combinations of values 
as shown in Table~\ref{tab:simulation_parameters}.
For each parameter setting, 
we generate $2$ \richel{Will need to calculate how many replicates we can do} 
independent reconstructed trees of the same crown age.

@@

From each simulated 'true' MBD tree, we measure the impact of
ignoring the more complex and non-standard MBD tree prior in
Bayesian phylogentic inference.
We do so by assuming a simpler and standard BD tree prior,
and compare the 'true' tree with the inferred trees.
To do so, we use the \verb;pirouette; R package \citep{pirouette}.

In brief: \verb;pirouette; starts from 
a 'true' starting phylogeny (in our case: the simulated MBD tree), 
from which a DNA sequence alignment is simulated. 
From each sequence alignment, a Bayesian 
inference is run (in our case: with assuming a BD tree prior), 
to obtain a posterior distribution of jointly-estimated trees and model parameter estimates.
By comparing the true tree and the posterior trees, 
an inference error distribution is generated.
We use the twinning option available in \verb;pirouette;
that allows to quantify the impact of assuming a wrong
tree prior in a Bayesian inference.

In our experiment, the alignments are 1000 nucleotides in length, 
with a known root sequence of four 250 mono-nucleotide blocks, 
generated using the simplest nucleotide substitution model (JC69) 
and clock model (strict), with a mutation rate of $\frac{1}{2}\cdot c$,
in which $c$ is the crown age. 
With this mutation rate, each nucleotide has a 50\% chance
to mutate (both silently and non-silently) from the ancestral root sequence 
to any of the contemporary species' sequences at the tips.

For the Bayesian inference, we assume a generative model of 
a JC69 site model, a strict clock model and a BD tree prior. 
Additionally, we use an MRCA prior equal to the crown age with a normal distribution 
of width $\sigma = 0.01$. We pick an MCMC setup of $10^6$ states
with a sampling interval of once per $10^3$ states. Of the resulting
$10^3$ states, we discard a burn-in of $10\%$.

To verify that the generative model (with a BD tree prior) is
indeed the most appropriate standard model, we compare it to
39 other candidate inference models. These candidate models 
are all other combinations of four tree priors, 
two clock models and five speciation models, 
excluding the one combination used as a generative model.
We let the best candidate model also generate an error distribution,
which is expected to have a similar or high median.
For models comparison, \verb;pirouette; uses a nested sampling approach
to estimate marginal likelihoods,
as described in [\cite{maturana2018model}], 
using the \verb;NS; BEAST2 package. We setup the MCMC for the nested sampling
using its default values, which are a relative error $\epsilon$ of $10^-12$, 
1 particle, a sub-chain sampling interval of $5.10^3$ states with
a maximum chain length of $10^6$ states.

To compare the true tree with any of the posterior trees,
we use the absolute nLTT statistic \citet{janzen2015},
which will result in an error distribution with values
ranging from zero (when the inferred tree is identical 
to the true tree) to a maximum of one (trees are completely different).

For the twinning, we let a birth-death process generate one random tree,
so that the twin inference can be done with the correct phylogenetic
model of a JC69 site model, strict clock model and a BD tree prior.
